Death By Dropdown?

### 2025-03-24 conference-materials/2025_shiny-conf/slides.qmd
Revamps slides for BID framework visualization
- Enhances presentation by introducing more visuals and focusing on real-world applicability of the BID framework.
- Introduces `{bidux}` as a developer toolkit to assist in applying BID principles to Shiny applications, showcasing its features and usage.
- Simplifies and consolidates content to improve clarity and engagement during the presentation.

::: {.fragment}
<br>
::: {.r-fit-text}
We can guide users effortlessly to insights
:::
:::

@@

@@

::: {.notes}
- BID doesn't reinvent the wheel but packages existing knowledge into a structured process
- Makes bias anticipation an explicit step for designers
- Brings together multiple disciplines to create a comprehensive framework
:::

## Stage 1: Notice the Problem {background-color="#FFE6E6"}

:::: {.columns}
::: {.column width="60%"}
### Detect high cognitive friction points

- **Cognitive Load Theory** (Sweller, 1988)
- **Hick's Law** (Hick, 1952) 
- **Visual Hierarchies** (Tufte, 1983)
:::

::: {.column width="40%"}
![](https://media.giphy.com/media/v1.Y2lkPTc5MGI3NjExNzkyZWQ1ZWJkNDk0NzFjMDMyNmE4ZjBkMWZmYmQ0Mjk3M2FmZGZjNCZlcD12MV9pbnRlcm5hbF9naWZ5X2dpZiZjdD1n/3o7TKCNxTFSY5psUUw/giphy.gif)
:::
::::

::: {.fragment}
### Actionable Example:
```r
# Instead of:
selectInput("var1", "Select Variable:", choices = names(data))
selectInput("var2", "Select Variable:", choices = names(data))
selectInput("var3", "Select Variable:", choices = names(data))

# Consider:
radioButtons("primaryVar", "Primary Metric:", 
             choices = c("Revenue", "Conversion", "Engagement"))
```
:::

## Stage 2: Interpret the User's Need {background-color="#E6F0E6"}

:::: {.columns}
::: {.column width="50%"}
### Understand the story users need

- **Data Storytelling Framework** (Matei & Hunter, 2021)
- **Processing Fluency** (Alter & Oppenheimer, 2009)
- **Emotion & Fluency Effects** (Song & Schwarz, 2009)
:::

::: {.column width="50%"}
### Ask key questions:
- What insight are they trying to gain?
- What decisions need to be made?
- What is the *simplest* way to present this information?
:::
::::

::: {.fragment}
### Actionable Example:
```r
# Instead of showing all data at once, guide the user:
fluidRow(
  box(title = "Key Business Question", 
      "Are we on track to meet Q2 targets?",
      status = "primary"),
  box(valueBoxOutput("mainKPI"))
)
```
:::

## Stage 3: Structure the Dashboard {background-color="#E6F0F0"}

### Build with intentional structure

- **Principle of Proximity** (Gestalt)
- **Dual-Processing Theory** (System 1 & 2)
- **Default Effect** (Johnson & Goldstein, 2003)
- **Aesthetic-Usability Effect** (Norman, 2002)

::: {.fragment}
### In Practice:
- Group related elements visually
- Provide quick summary cues AND detailed views
- Use smart defaults to reduce friction
- Ensure visual consistency and appeal
:::

## Stage 3: Structure In Action {background-color="#E6F0F0"}

```r
library(reactable)

# Applying Gestalt principles and defaults
reactable(
  data,
  defaultPageSize = 5,
  defaultSorted = list(revenue = "desc"),
  groupBy = "department",
  columns = list(
    revenue = colDef(
      format = colFormat(prefix = "$", separators = TRUE),
      style = function(value) {
        # System 1 color cues
        if (value > 1000000) {
          color <- "#008000"  # Green for good
        } else if (value < 500000) {
          color <- "#B22222"  # Red for concerning
        } else {
          color <- "#000000"  # Default
        }
        list(color = color, fontWeight = "bold")
      }
    )
  )
)
```

## Stage 4: Anticipate User Behavior {background-color="#F0E6F0"}

:::: {.columns}
::: {.column width="50%"}
### Consider biases & heuristics

- **Anchoring Effect** (Tversky & Kahneman, 1974)
- **Framing & Loss Aversion** (Tversky & Kahneman, 1981)
- **Confirmation Bias** (Nickerson, 1998)
- **Belief Perseverance** (Ross & Lepper, 1980)
:::

::: {.column width="50%"}
### Mitigation strategies:
- Provide context for all numbers
- Frame data positively when appropriate
- Show alternative views/analyses
- Include "disconfirming" visualizations
:::
::::

## Stage 4: Anticipate In Action {background-color="#F0E6F0"}

```r
# Dealing with anchoring and framing
tabsetPanel(
  tabPanel("Performance", 
    fluidRow(
      box(
        title = "Sales Performance",
        plotOutput("salesPlot")
      ),
      box(
        title = "Context",
        # Provide context to prevent anchoring
        "Industry average: $1.2M",
        "Company target: $1.5M",
        "Historical range: $0.9M - $1.8M"
      )
    ),
    # Alternative view to combat confirmation bias
    checkboxInput("showAlternative", "Show alternative analysis")
  ),
  tabPanel("What If?", 
    # Encourage exploration of contrary hypotheses
    sliderInput("scenario", "If market drops by:", min = 0, max = 30, value = 10),
    plotOutput("scenarioPlot")
  )
)
```

## Stage 5: Validate & Empower {background-color="#F0F0E6"}

:::: {.columns}
::: {.column width="50%"}
### Close the loop with clarity

- **Peak-End Rule** (Kahneman et al., 1993)
- **Beautiful-Is-Good Stereotype** (Dion et al., 1972)
- **Cooperation & Coordination** (Tindale & Kluwe, 2015)
:::

::: {.column width="50%"}
### Implementation:
- Strong, clear summary sections
- Visually appealing final views
- Collaborative features for teams
- Exportable insights
:::
::::

## Stage 5: Validate In Action {background-color="#F0F0E6"}

```r
# Peak-End Rule and Beautiful-Is-Good implementation
fluidRow(
  box(
    title = "Key Insights",
    status = "primary", 
    solidHeader = TRUE,
    width = 12,
    # Provide clear summary
    HTML("<h3>3 Key Takeaways:</h3>
         <ol>
           <li>Revenue is up 12% against target</li>
           <li>Western region showing 3x growth</li>
           <li>Product X is underperforming by 8%</li>
         </ol>"),
    # Enable collaboration
    actionButton("share", "Share Report", icon = icon("share")),
    downloadButton("downloadReport", "Export PDF")
  )
)
```

## Bringing It All Together {background-color="#FFFFFF"}

:::: {.columns}
::: {.column width="50%"}
### Before BID:
- Information overload
- No clear story
- Cognitive friction
- Unused features
- Unclear next steps
:::

@@

## Practical Examples {background-color="#E6F0FF"}

### Death By Dropdown ➡️ Decision By Design

::: {.nonincremental}
- **Grouped Headers**: Organize complex data into logical hierarchies
- **Dynamic Drilldowns**: Start with big picture, then reveal details
- **Inverse Color Scales**: Draw attention to what matters most
- **Narrative Pathways**: Guide users through a data story
:::

## Implementation Demo {background-color="#F0F0F0"}

```r
library(reactable)
library(echarts4r)

# Before: Overwhelming dropdown experience
# selectInput("region", "Region:", choices = regions)
# selectInput("product", "Product:", choices = products)
# selectInput("timeframe", "Time:", choices = timeframes)
# selectInput("segment", "Segment:", choices = segments)
# selectInput("metric", "Metric:", choices = metrics)

# After: Dynamic exploration with narrative focus
reactable(
  data,
  groupBy = "category",
  defaultExpanded = TRUE,
  columns = list(
    value = colDef(
      name = "Performance",
      cell = function(value, index) {
        # Contextual visualization
        sparklinePlot <- sparkline(data$trend[index], 
                                   lineColor = ifelse(value > target, "#28a745", "#dc3545"))
        div(sparklinePlot, style = "width: 100px; height: 30px;")
      }
    )
  ),
  details = function(index) {
    # Drill-down with context
    htmltools::div(
      echarts4r::e_charts(detail_data[detail_data$id == data$id[index], ]) |>
        e_line(x = "date", y = "value") |>
        e_tooltip() |>
        e_title(paste("Detailed View:", data$name[index]))
    )
  }
)
```

## Key Takeaways {background-color="#4682B4" .center}

::: {.nonincremental}
1. **Notice** where users struggle with cognitive load
2. **Interpret** the story they need, not just the data they have
3. **Structure** dashboards to guide attention and reduce friction
4. **Anticipate** biases and heuristics that affect decisions
5. **Validate** with clear summaries and next steps
:::

## Resources {background-color="#F0F0F0"}

### Tools to implement BID principles:

::: {.nonincremental}
- **{reactable}**: Advanced tables with grouped headers and dynamic drill-downs
- **{echarts4r}**: Interactive visualizations with narrative capabilities
- **{bslib}**: Create cohesive, branded dashboards
- **{shinysurveys}**: Gather user feedback to refine your approach
:::

::: {.fragment}
### Learn more:
- Code examples: [github.com/eickhoff-winget/BID-examples](https://github.com)
- Documentation: [bid-framework.org](https://example.com)
:::

## Thank You! {background-color="#4682B4" .center}

@@

@eickhoff_ux | @jwinget_data

## References {.scrollable}

- Alter, A. L., & Oppenheimer, D. M. (2009). Uniting the tribes of fluency. *Personality and Social Psychology Review, 13*(3), 219-235.
- Dion, K., Berscheid, E., & Walster, E. (1972). What is beautiful is good. *Journal of Personality and Social Psychology, 24*(3), 285-290.
- Hick, W. E. (1952). On the rate of gain of information. *Quarterly Journal of Experimental Psychology, 4*(1), 11-26.
- Johnson, E. J., & Goldstein, D. (2003). Do defaults save lives? *Science, 302*(5649), 1338-1339.
- Kahneman, D., Fredrickson, B. L., Schreiber, C. A., & Redelmeier, D. A. (1993). When more pain is preferred to less: Adding a better end. *Psychological Science, 4*(6), 401-405.
- Lockton, D., Harrison, D., & Stanton, N. A. (2010). The Design with Intent Method: A design tool for influencing user behaviour. *Applied Ergonomics, 41*(3), 382-392.
- Matei, S. A., & Hunter, J. L. (2021). Data storytelling is not storytelling with data. *The Information Society, 37*(5), 310-325.
- Nickerson, R. S. (1998). Confirmation bias: A ubiquitous phenomenon in many guises. *Review of General Psychology, 2*(2), 175-220.
- Norman, D. A. (2002). *The Design of Everyday Things.* Basic Books.
- Oinas-Kukkonen, H., & Harjumaa, M. (2009). Persuasive systems design: Key issues, process model, and system features. Communications of the Association for Information Systems, 24(1), 28.
- Ross, L., & Lepper, M. (1980). The persistence of inflated causal judgments. Journal of Personality and Social Psychology, 39(3), 341-352.
- Slovic, P. (1987). Perception of risk. Science, 236(4799), 280-285.
- Song, H., & Schwarz, N. (2009). If it's hard to read, does that mean it's less true? Psychological Science, 20(9), 1117-1120.
- Sweller, J. (1988). Cognitive load during problem solving: Effects on learning. Cognitive Science, 12(2), 257-285.
- Tindale, R. S., & Kluwe, K. (2015). Decision making in groups and organizations. APA Handbook of Personality and Social Psychology, 2, 603-623.
- Tufte, E. R. (1983). The Visual Display of Quantitative Information. Graphics Press.
- Tversky, A., & Kahneman, D. (1974). Judgment under uncertainty: Heuristics and biases. Science, 185(4157), 1124-1131.
- Tversky, A., & Kahneman, D. (1979). Prospect theory: An analysis of decision under risk. Econometrica, 47(2), 263-291.
- Tversky, A., & Kahneman, D. (1981). The framing of decisions and the psychology of choice. Science, 211(4481), 453-458.
- Vessey, I. (1991). Cognitive fit: A theory-based analysis of the graphs versus tables literature. Decision Sciences, 22(2), 219-240.

>>>>

::: {.fragment}

::: {.r-fit-text}
We can guide users effortlessly to insights
:::
:::

@@

> _(placeholder for BID framework visual)_

@@

## Real-World Patterns, Not Just Theory {.center background-color="#FFFFFF"}

::: {.columns}
::: {.column width="50%"}
### Before BID:
- Information overload
- No clear story
- Cognitive friction
- Unused features
- Unclear next steps
:::

@@

## BID In Action {.center background-color="#F0F0F0"}

::: {.r-fit-text}
**Screenshots**: Before & After applying BID framework across stages
:::

> _(placeholder for visual comparison of UI examples)_

## Implementation Demo {.center background-color="#F9F9F9"}

::: {.r-fit-text}
Hands-on example using `{reactable}` + `{echarts4r}` to implement Stages 1–5
:::

> _(placeholder for staged code walkthrough or animated screen recording)_

## From Framework to Workflow {.center background-color="#E6F0FF"}

::: {.r-fit-text}
Meet `{bidux}`: Your BID Assistant for Shiny
:::

- 🧠 Five structured helpers: `bid_notice()` to `bid_validate()`
- 📚 Concept browser: `bid_concept("processing_fluency")`
- 📄 Built-in documentation + report generation
- 🔌 Works with `{rhino}` and your existing apps

> _(placeholder screenshot: console usage or UI screen of `{bidux}` helpers)_

## What bidux Looks Like {.background-color="#F9F9F9"}

```r
library(bidux)

ui <- fluidPage(
  bid_structure(
    layout = "dual_process",
    concepts = c("default_effect", "principle_of_proximity")
  )
)

ui <- bid_anticipate(
  ui,
  bias_mitigations = list(
    anchoring = bid_components$context_aware_reference("sales_baseline"),
    framing = bid_components$framing_toggle("performance_view", "Progress", "Gap")
  )
)
```

::: {.fragment}
More at: [github.com/bidux](https://github.com)
:::

## Key Takeaways {.center background-color="#4682B4"}

::: {.nonincremental}
1. **BID** helps reduce friction and improve decisions
2. **Reactable + echarts4r** bring BID to life
3. **`bidux`** supports you in applying BID at every stage
4. Build dashboards that guide — not overwhelm — your users
:::

## Thank You! {.center background-color="#4682B4"}

@@
